---
title: "2022 02 03 Bayesian Piecewise Exponential Models"
subtitle: ""
excerpt: ""
date: 2022-02-03T18:47:55-05:00
author: ""
draft: true
series:
tags:
categories:
layout: single # single or single-sidebar
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)
library(ggthemes)
library(patchwork)
```
## The Exponential Distribution 

<!-- The [Exponential Distribution](https://en.wikipedia.org/wiki/Exponential_distribution)  -->

An exponential survival model is considered the simplest [parametric survival model](https://data.princeton.edu/wws509/notes/c7.pdf). An exponential model assumes the risk of an event occurring remains constant over time. This results in the the hazard function
$$h(t) = \lambda$$
and the corresponding survival function given by
$$S(t) = \exp(-\lambda t) \text{ and } $$

```{r exponential_plot, fig.cap = "Hazard and Survival Function"}
time_seq <- seq(0, 120, by = 0.5)
lambda_fixed <- log(2) / 40

exp_plot_df <-
  tibble("t" = time_seq, "hazard" = lambda_fixed) %>%
  mutate("s_of_t" = pexp(q = t, rate = hazard, lower.tail = FALSE))

exp_surv_plot <-
  ggplot(exp_plot_df) +
  geom_line(
    aes(x = t, y = s_of_t),
    size = 1.5
  ) +
  theme_minimal() +
  labs(
    title = "Survival Function",
    x = "Time",
    y = "Prob. of Survival"
  )

exp_haz_plot <-
  ggplot(exp_plot_df) +
  geom_line(
    aes(x = t, y = hazard),
    size = 1.5
  ) +
  theme_minimal() +
  labs(
    title = "Survival Function",
    x = "Time",
    y = "Hazard"
  )
```

## Why use an Exponential Model

## Piecewise Exponential Model
### One Changepoint / Two Hazards

```{r two_hazard_pw}
time_seq <- seq(0, 120, by = 0.5)
lambda_1 <- log(2) / 40
lambda_2 <- log(2) / 15
changepoint <- 25

two_pwe_plot_df <-
  tibble("t" = time_seq) %>%
  mutate(
    "hazard" = if_else(t <= changepoint, lambda_1, lambda_2),
    "s_of_t" = pexp(q = t, rate = hazard, lower.tail = FALSE)
  )

two_pwe_surv <- ggplot(two_pwe_plot_df) +
  geom_line(
    aes(x = t, y = s_of_t),
    size = 1.5
  ) +
  theme_minimal() +
  labs(
    title = "Survival Function",
    x = "Time",
    y = "Prob. of Survival"
  )

two_pwe_haz <- ggplot(two_pwe_plot_df) +
  geom_line(
    aes(x = t, y = hazard),
    size = 1.5
  ) +
  theme_minimal() +
  labs(
    title = "Hazard Function",
    x = "Time",
    y = "Hazard"
  )

two_pwe_haz / two_pwe_surv

```

## Fitting a Piecewise Exponential with Stan

```{r echo = TRUE}
library(PWEALL) # Simulates piecewise exponential

# Simulate survival times -----------------------------------------------------
# ctrl_time = rpwe(nr = n_ctrl, rate = c(rate_1_ctrl, rate_2_ctrl), tchange = c(0,change_point))$r
# trt_time = rpwe(nr = n_trt, rate = c(rate_1_trt, rate_2_trt), tchange = c(0,change_point))$r
# 
# all_data = tibble(
#   "id" = seq(1, n_trt + n_ctrl),
#   "time" = c(ctrl_time, trt_time),
#   "trt" = c(rep(0, n_ctrl), rep(1, n_trt)),
#   "event" = 1 # everyone has an event, replace with censoring in actual data
# )

```

